---
title: "Priya-Mina: Analysis of Temperature and PM 2.5 levels in Tamil Nadu"
subtitle: "ESPM 157 Final Project"
date: "December 2023"
author: "Mina Burns and Priya Riley"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

Title: Analysis of Temperature and PM 2.5 levels in Tamil Nadu

Description of the project: For this project we will use data from Low Cost Sensors in habitations in rural Tamil Nadu to investigate relationships between temperature, humidity and PM 2.5 levels in the region. 

Data required and how it is obtained:
Data from low cost sensor network deployed in Tamil Nadu, districts Kallakurichi and Nagapattinam.
AAM-LASSI dataset. Ambient Air Monitoring of LPG At Scale in South India 
From Mina’s research with the Center for Occupational and Environmental Health 

Three questions / analysis tasks:
1) Look at connections between temperature and PM 2.5 in one village in India over a time period of about one year. 
2) Investigate relationship between humidity and PM 2.5 in one village in India over a time period of about one year. 
3) Compare the two districts, Kallakurichi and Nagapattinam (coastal and inland, respectively), for PM2.5 measurements. 

--------------------------------------------
<<<<<<< HEAD


# Preliminary EDA - Priya - using the AQM_Inventory.xlsx

=======
# Preliminary EDA - Priya - using the AQM_Inventory.xlsx

TN? Kallakurichi and Nagapattinam. KK? NP?those exist in the data set so ima say yes.

>>>>>>> 1e9f6b959a9ac1bf15774fa50a4b2bda3528901e
```{r setup, include=FALSE}
#Read in packages
library("readxl")
library("base") 
library("data.table")
library("datasets")
library("DT")
library("dplyr")
library("ggplot2")
library("graphics")
library("readxl")
library("tidyr")
```

```{r}
#read in metadata 
metadata <- read_excel("~/Desktop/STEER/AQM_Inventory.xlsx")
metadata
```

```{r}
#I only want the NP and KK locations, because we are looking at Kallakurichi and Nagapattinam.
#Also, I only want to keep rows where the current status is 'Working'.
filtered_metadata = filter(metadata, `Current Location` == "NP" | `Current Location` == "KK")
filtered_metadata = filter(filtered_metadata, `Current Status` == "Working")
filtered_metadata

```


#### Data visualizations using outdoor cleaned data 


```{r}
<<<<<<< HEAD
# read the csv into r. The csv covers ambient and household emissions data in Tamil Nadu, India.

daily_emissions_data = read.csv("~/Desktop/STEER/dailymeans_outdoor_HH_cleaned.csv") 
  #filter(cleaned_mina_data, #flags greater than 0?)

#filter where the qa_flags and da_flags are above 0, which means they have been flagged for an error.
cleaned_data = filter(daily_emissions_data, qa_flag == "0" | da_flag == "0")


!
=======
cleaned_mina_data = read.csv("~/Desktop/STEER/dailymeans_outdoor_HH_cleaned.csv")
cleaned_mina_data
# of course, later on I can change the names to not call it mina_data to be more professinonal. But for now it is helpful to me!
>>>>>>> 1e9f6b959a9ac1bf15774fa50a4b2bda3528901e
```

```{r}
#real quick, let's make sure that all variables are either humidity (rh), temperature (temp), or pm2.5
unique(cleaned_mina_data$variable)
```

Great! Now lets move on to create one table for district KK and one for NP.

```{r}
#lets get one table for district KK and one for NP. First, let's do KK. 
<<<<<<< HEAD
KK_mina_cleaned = filter(cleaned_mina_data, district == "KK") 
=======
KK_mina_cleaned = filter(cleaned_mina_data, district == "KK")
>>>>>>> 1e9f6b959a9ac1bf15774fa50a4b2bda3528901e
KK_mina_cleaned
```

```{r}
<<<<<<< HEAD
#Now let's get a table for district NP.
=======
#Now let's get a table for distict NP.
>>>>>>> 1e9f6b959a9ac1bf15774fa50a4b2bda3528901e
NP_mina_cleaned = filter(cleaned_mina_data, district == "NP")
NP_mina_cleaned
```

Great - all preliminary tables are set up. Now we can explore the data and our questions. Let's begin by exploring district KK.

```{r}
<<<<<<< HEAD
#For our question, we only need to look at the date, district, variable, mean_of_medians. 
KK_selected_data = select(KK_mina_cleaned, c("date", "district", "variable", "mean_of_medians"))
KK_selected_data


##
```
=======
#For our question, we only need to look at the date, district, variable, mean_of_medians, qa_flag, and da_flag.
KK_selected_data = select(KK_mina_cleaned, c("date", "district", "variable", "mean_of_medians", "qa_flag", "da_flag"))
KK_selected_data
```
#### Uh oh! it looks like all this data for district KK is in the span of one month. That will not allow us to answer our 3rd research question ! (at least, this alone will not be able to answer it. maybe these plus the missing rds to create a merged table, would allow us to answer the third question. )
>>>>>>> 1e9f6b959a9ac1bf15774fa50a4b2bda3528901e

### Moving on to district NP for the first 2 questions.


```{r}
#For our first two questions, we only need to look at the date, district, variable, and mean_of_medians.
NP_selected_data = select(NP_mina_cleaned, c("date", "district", "variable", "mean_of_medians"))
NP_selected_data
```

<<<<<<< HEAD
=======
Above, we see that this data goes only from october to december, which is two months. Thus this limited data set also will not answer our first two questions, but I will do a few graphs regardless. Once we get the full data figured out then I can make graphs for it to fully work. 

>>>>>>> 1e9f6b959a9ac1bf15774fa50a4b2bda3528901e
1) Look at connections between temperature and PM 2.5 in one village in India over a time period of about one year. 

In this case the chosen village is NP: Nagapattinam.
```{r}

<<<<<<< HEAD
#lets make columns for temp and pm2.5
=======
#lets make three columns, for pm2.5, rh, and temp. --> thats hard to do! eek

#ok instead lets make columns for temp and pm2.5
>>>>>>> 1e9f6b959a9ac1bf15774fa50a4b2bda3528901e

filtered_pm2.5 = NP_selected_data %>%
  filter(variable == "pm2.5") %>%
  arrange(date, district) %>%
  mutate(pm25_values = mean_of_medians)
#filtered_pm2.5

filtered_temp = NP_selected_data %>%
  filter(variable == "temp") %>%
  arrange(date, district) %>%
  mutate(temp_values = mean_of_medians)
#filtered_temp
```
When I look at the number of inputs per date, I see that they have different numbers. Therefore I need to make them into an average value per date. 

```{r}
average_NP_pm2.5 = filtered_pm2.5 %>%
  group_by(date) %>%
  summarise(avg_pm2.5 = mean(pm25_values))

average_NP_pm2.5
```

```{r}
average_NP_temp = filtered_temp %>%
  group_by(date) %>%
  summarise(avg_temp = mean(temp_values))

average_NP_temp
```

Awesome. Now I can make some plots with these averages.

```{r}
ggplot() + 
  geom_point(data = average_NP_pm2.5, aes(x=date, y = avg_pm2.5)) +
  geom_point(data = average_NP_temp, aes(x=date, y=avg_temp), color = "blue") +
  ggtitle("Avg temperature and PM 2.5 concentrations per day") + 
  labs(x= "Date", y = "Mean Values") +
  scale_x_discrete(name = "Date", breaks = c(0:10))

```

This isn't super helpful to me. Lets break it down and see if that helps.


```{r}
avg_temp_plot = ggplot() + 
  geom_point(data = average_NP_temp, aes(x=date, y=avg_temp), color = "coral") +
  ggtitle("Avg temperature values per day") + 
  labs(x= "Date", y = "Mean Values") +
  scale_x_discrete(name = "Date", breaks = c(0:10)) 

avg_pm25_plot = ggplot() + 
  geom_point(data = average_NP_pm2.5, aes(x=date, y = avg_pm2.5), color = "lightblue") +
  ggtitle("Avg PM 2.5 concentrations per day") + 
  labs(x= "Date", y = "Mean Values") +
  scale_x_discrete(name = "Date", breaks = c(0:10)) 

avg_temp_plot
avg_pm25_plot
```

```{r}
#the graphs next to each other
require(gridExtra)
grid.arrange(avg_temp_plot, avg_pm25_plot, ncol=2)

```

We can see there are somewhat similar patterns. It looks like temp increases right after pm2.5 increases.

```{r}
avg_temp_plot_col = ggplot() + 
  geom_col(data = average_NP_temp, aes(x=date, y=avg_temp), color = "coral") +
  ggtitle("Avg temperature values per day") + 
  labs(x= "Date", y = "Mean Values") +
  scale_x_discrete(name = "Date", breaks = c(0:10)) 

avg_pm25_plot_col = ggplot() + 
  geom_col(data = average_NP_pm2.5, aes(x=date, y = avg_pm2.5), color = "lightblue") +
  ggtitle("Avg PM 2.5 concentrations per day") + 
  labs(x= "Date", y = "Mean Values") +
  scale_x_discrete(name = "Date", breaks = c(0:10)) 

require(gridExtra)
grid.arrange(avg_temp_plot_col, avg_pm25_plot_col, ncol=2)

###it would be nice to have them as lines instead, on the same plot. work on that tmrw
```



2) Investigate relationship between humidity and PM 2.5 in one village in India over a time period of about one year. 
```{r}

```

```{r}

```